MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-07-02 main/MSstatsPTM_Main.tex
Implementation

$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a series of R/Bioconductor packages designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. Models in these packages takes as input a list of log$_2$-intensities $y_{ijk}$. For each protein, the feature log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}. These summarized values are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}. The specific model instances depend on the experimental design and data acquisition. For example, for an experiment with \todo{group comparison or time course design?} $MSstatsTMT$ fits the following model \todo{OV: could we keep the notation as consistent as possible across this section? Above conditions had subscripts $i$ and bio reps had subscripts $j$. Could we keep a consistent notation below (or change the notation elsewhere in the manuscript?)} \todo{Should we maybe give a model for label-free, to stay consistent with the example in figure 2? Just use $I$ and $J$ etc for the indices...}
\begin{eqnarray}
& y_{mtcb} = \mu + Mixture_m + TechRep(Mixture)_{t(m)} + Condition_c + Subject_{mcb} + \epsilon_{mtcb}, & \\ 
& \text{where }  Mixture_m  \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_M),  TechRep(Mixture) \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_T), &  \nonumber \\
& \sum\limits_{c=1}^{C} Condition_c = 0,\ \todo{Subject \stackrel{iid}{\sim}\mathcal{N}(0,?)}\  \epsilon_{mtcb} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
\end{eqnarray}
Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ \todo{this has not been defined, Define here, using the notation above?} and its standard error.
Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}. 
Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.

@@

\subsection*{Detection of changes in PTMs}
\label{sec:adjustment_summary}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Implementation}

The proposed methods are implemented in the open source R package $MSstatsPTM$. An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$. Conversion is done separately for the modified peptide and global profiling runs. If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance. This will reduce our null hypotheses to Equation \ref{eq:conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.

After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package summarizes features which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package can optionally impute missing values using an AFT model~\cite{Wei:1992}. Missing value imputation should be performed with care when applied to PTM data. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can cause the missing value imputation to be unstable.

@@

Beyond the core functionality, the package also includes functions for plotting the results. These include summarization plots to help with quality assurance and identifying sources of variation (such as quality control plots and profile plots), as well as modeling plots (such as volcano plots and heatmaps), which evaluate the fold change between conditions and peptides.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The proposed approach corrected for high FDR in computer simulations}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The proposed approach outperformed existing methods in simulations with missing values and low features}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In a label-free benchmark experiment the proposed approach corrected for bias and outperformed existing methods}

@@

\subsubsection*{Proposed approach removed confounding in two mixture TMT experiment}

@@

\subsubsection*{In label-free experiment without a separate global profiling run, the proposed approach could eliminate the bias due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}

>>>>

$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a series of R/Bioconductor packages designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. Models in these packages takes as input a list of log$_2$-intensities $y_{ijk}$. For each protein, the feature log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}. These summarized values are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}. The specific model instances depend on the experimental design and data acquisition. For example, for an experiment with \todo{group comparison or time course design?} $MSstatsTMT$ fits the following model \todo{OV: could we keep the notation as consistent as possible across this section? Above conditions had subscripts $i$ and bio reps had subscripts $j$. Could we keep a consistent notation below (or change the notation elsewhere in the manuscript?)} \todo{Should we maybe give a model for label-free, to stay consistent with the example in figure 2? Just use $I$ and $J$ etc for the indices...}
\begin{eqnarray}
& y_{mtcb} = \mu + Mixture_m + TechRep(Mixture)_{t(m)} + Condition_c + Subject_{mcb} + \epsilon_{mtcb}, & \\ 
& \text{where }  Mixture_m  \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_M),  TechRep(Mixture) \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_T), &  \nonumber \\
& \sum\limits_{c=1}^{C} Condition_c = 0,\ \todo{Subject \stackrel{iid}{\sim}\mathcal{N}(0,?)}\  \epsilon_{mtcb} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
\end{eqnarray}
Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ \todo{this has not been defined, Define here, using the notation above?} and its standard error.
Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}. 
\todo{OV: maybe elaborate on that?} Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.

@@

\subsection*{Statistical methods in MSstatsPTM}
\label{sec:adjustment_summary}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Implementation of MSstatsPTM}

\todo{OV: I feel that some parts of this section are either specific to MSstats/MSstatsTMT, or a bit redundant with the methods. For the parts that we do not implement anew but reuse, could you move them to the 'Existing methods'/MSstats section? Here let us focus only on things that we had to develop from scratch for this project. If we implement the methods, we do not need to re-describe the methods}

\todo{OV: could you make this section more specific to PTM converters? E.g., I remember that the input should be different, there are some parts relevant to the localization of PTMs. This was not part of the base MSstats. Could you emphasize that?} The proposed methods are implemented in the open source R package $MSstatsPTM$. An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$. Conversion is done separately for the modified peptide and global profiling runs. If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance. This amounts to testing the null hypothesis in~\eqref{conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.

\todo{OV: this is the part that looks redundant with the MSstats section. In the MSstats section, do not repeat the methods, just focus on the specifics of the implementation. Relavant parts would be that we fit simplified models if the base model is not applicable. In this section,  could you just say that we do the same as MSstats (including fitting simplified models). But there are some caveats, e.g. assumptions of missing values due to low abundance are not necessarily appropriate here so we recommend to not use that option? Also specialized sample size calculations} After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package summarizes features which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package can optionally impute missing values using an AFT model~\cite{Wei:1992}. Missing value imputation should be performed with care when applied to PTM data. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can cause the missing value imputation to be unstable.

@@

\todo{OV: similarly, some of this belongs to MSstats. Could you focus on project-specific implementation. E.g., we separately visualize modified and unmodified peptides. (Btw, sequence-specific visualization such as the one you did for MSstatsLiP would be awesome if there is a chance)} Beyond the core functionality, the package also includes functions for plotting the results. These include summarization plots to help with quality assurance and identifying sources of variation (such as quality control plots and profile plots), as well as modeling plots (such as volcano plots and heatmaps), which evaluate the fold change between conditions and peptides.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{MSstatsPTM corrected for high FDR in computer simulations}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{MSstatsPTM outperformed existing methods in simulations with missing values and low features}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In a label-free benchmark experiment MSstatsPTM corrected for bias and outperformed existing methods}

@@

\subsubsection*{MSstatsPTM removed confounding in two mixture TMT experiment}

@@

\subsubsection*{In label-free experiment without a separate global profiling run, MSstatsPTM eliminated the bias due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}
